---
title: "Kings Project R Code"
author: "Ian Turner"
date: "2025-10-16"
output: pdf_document
---
# Setup: Libraries & Global Options
```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(jsonlite)
library(dplyr)
library(lubridate)
library(MASS)
library(pROC)
library(caret)
library(ggplot2)
library(scatterplot3d)
library(tidyr)
library(shiny)
library(DT)
library(plotly)

```

# Parse JSON CSV & Load Data
```{r}

convert_json_to_csv <- function(in_file, out_file) {
  txt <- readLines(in_file, warn = FALSE, encoding = "UTF-8")
  txt <- paste(txt, collapse = "\n")
  objs <- unlist(regmatches(txt, gregexpr("\\{[\\s\\S]*?\\}", txt, perl = TRUE)))
  if (length(objs) == 0) stop("No JSON objects found.")
  json_arr <- paste0("[", paste(objs, collapse = ","), "]")
  json_arr <- gsub('(:)\\s*-\\s*(?=[,}])', '\\1 null', json_arr, perl = TRUE)
  if (!jsonlite::validate(json_arr)) stop("JSON invalid after cleanup.")
  dat <- fromJSON(json_arr, flatten = TRUE)
  write.csv(dat, out_file, row.names = FALSE)
  cat(sprintf("%s → %s (%d rows, %d cols)\n", in_file, out_file, nrow(dat), ncol(dat)))
}

convert_json_to_csv("nba_box_player_season.json", "nba_data.csv")
nba_data <- read.csv("nba_data.csv", header = TRUE)

convert_json_to_csv("international_box_player_season.json",
                    "international_data.csv")
international_data <- read.csv("international_data.csv")

convert_json_to_csv("player.json", "player_data.csv")
player_data <- read.csv("player_data.csv")

```





# Normalize Player Names (Title Case)
```{r}

# Capitalize each name properly
international_data$first_name <- tools::toTitleCase(tolower(international_data$first_name))
international_data$last_name  <- tools::toTitleCase(tolower(international_data$last_name))

player_data$first_name <- tools::toTitleCase(tolower(player_data$first_name))
player_data$last_name  <- tools::toTitleCase(tolower(player_data$last_name))

```




# Aggregate Seasons and Player Totals & Weighted Averages
```{r}
wmean_safe <- function(x, w) {
  ok <- is.finite(x) & is.finite(w) & w > 0
  if (!any(ok)) return(NA_real_)
  weighted.mean(x[ok], w[ok])
}

# totals to SUM
totals_cols <- c(
  "games","starts","minutes","points",
  "two_points_made","two_points_attempted",
  "three_points_made","three_points_attempted",
  "free_throws_made","free_throws_attempted",
  "blocked_shot_attempts","offensive_rebounds","defensive_rebounds",
  "assists","screen_assists","turnovers","steals","deflections",
  "loose_balls_recovered","blocked_shots","personal_fouls","personal_fouls_drawn",
  "offensive_fouls","charges_drawn","technical_fouls","flagrant_fouls","ejections",
  "points_off_turnovers","points_in_paint","second_chance_points","fast_break_points", 
  "possessions", "estimated_possessions", "team_possessions"
)

wa_cols <- names(international_data)[41:52]

player_intl_totals <- international_data %>%
  group_by(first_name, last_name) %>%
  mutate(w_games = games) %>%                       # <- preserve per-season weights
  summarise(
    across(all_of(totals_cols), ~ sum(.x, na.rm = TRUE)),
    across(all_of(wa_cols), ~ wmean_safe(.x, w_games), .names = "{.col}_wa"),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2))) %>%
  arrange(last_name, first_name)

```



# Add Age & NBA Experience (as of 2021-12-31)
```{r}
# helper so joins actually match
fmt <- function(x) tools::toTitleCase(tolower(x))

# --- standardize names in all three sources ---
player_intl_totals_std <- player_intl_totals %>%
  mutate(first_name = fmt(first_name),
         last_name  = fmt(last_name))

player_data_std <- player_data %>%
  mutate(first_name = fmt(first_name),
         last_name  = fmt(last_name),
         birth_date = ymd(birth_date))

nba_players_lookup <- nba_data %>%
  mutate(first_name = fmt(first_name),
         last_name  = fmt(last_name)) %>%
  distinct(first_name, last_name) %>%
  mutate(nba_experience = TRUE)

# Freeze time at end of 2021 
ref_date <- ymd("2021-12-31")

player_intl_totals_final <- player_intl_totals_std %>%
  left_join(
    player_data_std %>% dplyr::select(first_name, last_name, birth_date),
    by = c("first_name", "last_name")
  ) %>%
  mutate(
    birth_date = as_date(birth_date),
    age = if_else(
      !is.na(birth_date),
      floor(time_length(interval(birth_date, ref_date), "years")),
      NA_integer_
    )
  ) %>%
  relocate(age, .after = last_name) %>%
  left_join(nba_players_lookup, by = c("first_name", "last_name")) %>%
  mutate(
    nba_experience = coalesce(nba_experience, FALSE) %>% as.logical()
  ) %>%
  relocate(nba_experience, .after = age)

```



# Convert Totals to Per-Game Rates
```{r}
intl_per_game <- player_intl_totals_final %>%
  mutate(across(
    c(minutes, points, two_points_made, two_points_attempted,
      three_points_made, three_points_attempted, free_throws_made,
      free_throws_attempted, blocked_shot_attempts, offensive_rebounds,
      defensive_rebounds, assists, screen_assists, turnovers, steals,
      deflections, loose_balls_recovered, blocked_shots, personal_fouls,
      personal_fouls_drawn, offensive_fouls, charges_drawn, technical_fouls,
      flagrant_fouls, ejections, points_off_turnovers,
      second_chance_points, fast_break_points, possessions, estimated_possessions, 
      team_possessions),
    ~ if_else(games > 0, .x / games, NA_real_)
  )) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

```

# Attach Most-Recent Season (Recency Key)
```{r}
recent_season <- international_data %>%
  group_by(first_name, last_name) %>%
  summarise(most_recent_year = max(season, na.rm = TRUE), .groups = "drop")

intl_per_game <- intl_per_game %>%
  left_join(recent_season, by = c("first_name", "last_name")) %>%
  relocate(most_recent_year, .after = nba_experience)
```


# Export Master Table (intl_per_game.csv)
```{r}
intl_per_game
write.csv(intl_per_game, "full_international_data.csv", row.names = FALSE)
```




# Logistic Regression: Train/Test, Stepwise AIC, Evaluate, Score Full Dataset
```{r}
set.seed(42)

# ---- 1) Data prep ----

df <- intl_per_game %>%
  mutate(nba_experience = as.numeric(nba_experience == TRUE)) %>%
  dplyr::select(nba_experience, 3:22, 37:51, -20, -5)  


# Replace NAs with column means
df <- df %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.x), mean(.x, na.rm = TRUE), .x)))

# ---- 2) Train/test split ----
set.seed(42)
train_index <- createDataPartition(df$nba_experience, p = 0.8, list = FALSE)
train <- df[train_index, ]
test  <- df[-train_index, ]

# ---- 3) Logistic regression + stepwise variable selection ----
base_model <- glm(nba_experience ~ ., data = train, family = binomial)

# AIC-based stepwise selection
step_model <- stepAIC(base_model, direction = "both", trace = FALSE)
summary(step_model)

# ---- 4) Predictions and NBA probability scores ----
test$nba_prob <- predict(step_model, newdata = test, type = "response")

# ---- 5) Model evaluation ----
roc_obj <- roc(test$nba_experience, test$nba_prob)
auc_val <- auc(roc_obj)
cat("AUC =", auc_val, "\n")

# Classification at 0.5 threshold
test$predicted_class <- ifelse(test$nba_prob >= 0.5, 1, 0)

confusion <- confusionMatrix(
  factor(test$predicted_class),
  factor(test$nba_experience),
  positive = "1"
)
print(confusion)

# ---- 7) Add NBA probability score to full dataset ----
intl_per_game <- intl_per_game %>%
  mutate(
    nba_probability = predict(step_model, newdata = df, type = "response")
  )

```


# High-Prob Non-NBA Candidates (False Positives)
```{r}
# Top non-NBA players by predicted probability
false_positives <- intl_per_game %>%
  filter(nba_experience == 0, most_recent_year >= 2019) %>%                 # not actually in NBA
  arrange(desc(nba_probability)) %>%              # sort by model score
  dplyr::select(first_name, last_name, age, nba_probability, everything())

false_positives

```

# Plot: NBA Probability Distribution (Threshold = 0.5)
```{r}
ggplot(intl_per_game, aes(x = nba_probability)) +
  geom_histogram(
    bins = 30,
    fill = "steelblue",
    color = "white",
    alpha = 0.8
  ) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "red", linewidth = 1) +
  labs(
    title = "Distribution of NBA Probability Scores",
    x = "Predicted NBA Probability",
    y = "Number of Players"
  ) +
  annotate("text", x = 0.52, y = Inf, label = "0.5 Threshold", vjust = -1, hjust = 0, color = "red", size = 3.5) +
  theme_minimal(base_size = 13)

```


# Standardize Features (z-Scores)
```{r}
intl_scaled <- intl_per_game %>%
  mutate(across(
    c(points, usage_percentage_wa, true_shooting_percentage_wa,
      three_points_made, three_point_attempt_rate_wa,
      assists, assist_percentage_wa, turnovers,
      offensive_rebounding_percentage_wa, defensive_rebounding_percentage_wa, total_rebounding_percentage_wa,
      blocked_shots, block_percentage_wa, defensive_rebounds),
    ~ scale(.) %>% as.numeric()
  ))

```


# Archetype Coefficients: Scorer (Points + TS%)
```{r}
model_interact <- lm(
  internal_box_plus_minus_wa ~ scale(points) +scale(true_shooting_percentage_wa),
  data = intl_per_game
)

summary(model_interact)

```



# Compute Scorer Score & Percentile
```{r}
intl_per_game <- intl_per_game %>%
  mutate(
    scorer_score = 1.51 * as.numeric(scale(points)) +
                   3.51 * as.numeric(scale(true_shooting_percentage_wa)),
    
    scorer_percentile = 100 * round(percent_rank(scorer_score), 3)
  )

```

# Archetype Coefficients: Shot Blocker (Blocks - Fouls)
```{r}
model_interact <- lm(
  internal_box_plus_minus_wa ~ scale(blocked_shots) + scale(-personal_fouls),
  data = intl_per_game
)

summary(model_interact)

```


# Archetype Coefficients: Shot Blocker (Blocks - Fouls)
```{r}

intl_per_game <- intl_per_game %>%
  mutate(
    shot_blocker_score =
      as.numeric(scale(blocked_shots))       * coef(model_interact)[["scale(blocked_shots)"]] +
      as.numeric(scale(-personal_fouls))      * coef(model_interact)[["scale(-personal_fouls)"]],
    shot_blocker_percentile = 100 * round(percent_rank(shot_blocker_score), 3)
  )


```




# Archetype Coefficients: Rebounder (OREB + DREB)
```{r}
model_rebounder_pct <- lm(
  internal_box_plus_minus_wa ~ 
    scale(offensive_rebounds) + 
    scale(defensive_rebounds),
  data = intl_per_game
)
summary(model_rebounder_pct)


intl_per_game <- intl_per_game %>%
  mutate(
    rebounder_score = 0.79 * scale(offensive_rebounds) +
                      2.16 * scale(defensive_rebounds),
    rebounder_percentile = 100 * round(percent_rank(rebounder_score), 3)
  )

```




# Archetype Coefficients: Facilitator (AST - TOV%) & Score
```{r}
intl_per_game <- intl_per_game %>%
  mutate(
    assist_to_turnover_ratio = ifelse(turnovers > 0, assists / turnovers, NA)
  )

model_facilitator <- lm(
  internal_box_plus_minus_wa ~ 
    scale(assists) +
    scale(turnover_percentage_wa),
  data = intl_per_game
)
summary(model_facilitator)

intl_per_game <- intl_per_game %>%
  mutate(
    facilitator_score =
      1.84 * as.numeric(scale(assists)) -     # from model coefficient
      2.27 * as.numeric(scale(turnover_percentage_wa)),  # inverse relationship
    facilitator_percentile = 100 * round(percent_rank(facilitator_score), 3)
  )

```


# Archetype Coefficients: Floor Spacer (3PM + TS%) & Score
```{r}

intl_per_game <- intl_per_game %>%
  mutate(
    three_point_percentage = ifelse(three_points_attempted > 0,
                                    three_points_made / three_points_attempted, NA),
    
    floor_spacer_score = 
      1.25* as.numeric(scale(three_points_made)) +
      3 * as.numeric(scale(true_shooting_percentage_wa)),
    
    floor_spacer_percentile = 100 * round(percent_rank(floor_spacer_score), 3)
  )
model_3p <- lm(
  internal_box_plus_minus_wa ~ 
    scale(three_points_made) +  scale(true_shooting_percentage_wa),
  data = intl_per_game
)
summary(model_3p)
```




# Archetype Coefficients: Post Scorer (2P%, 2PA, FT%, FTA, PF Drawn)
```{r}
# Create efficiency variables
intl_per_game <- intl_per_game %>%
  mutate(
    two_point_percentage = ifelse(two_points_attempted > 0,
                                  two_points_made / two_points_attempted, NA),
    free_throw_percentage = ifelse(free_throws_attempted > 0,
                                   free_throws_made / free_throws_attempted, NA)
  )

# Build the linear model
model_post_scorer_final <- lm(
  internal_box_plus_minus_wa ~ 
    scale(two_point_percentage) +
    scale(two_points_attempted) +
    scale(free_throws_attempted) +
    scale(free_throw_percentage) +
    scale(personal_fouls_drawn),
  data = intl_per_game
)

summary(model_post_scorer_final)

```




# Compute Post Scorer Score & Percentile
```{r}
intl_per_game <- intl_per_game %>%
  mutate(
    post_scorer_score =
      2.17 * as.numeric(scale(two_point_percentage)) +
      0.24 * as.numeric(scale(two_points_attempted)) +
      0.63 * as.numeric(scale(free_throw_percentage)) +
      0.99 * as.numeric(scale(personal_fouls_drawn)),

    post_scorer_percentile = 100 * round(percent_rank(post_scorer_score), 3)
  )

```




# Tiering & Top-3 Archetype Labels per Player
```{r}
get_tier <- function(x) {
  case_when(
    x >= 95 ~ "Elite",
    x >= 85 ~ "Great",
    x >= 70 ~ "Good",
    x >= 50 ~ "Above Average",
    x >= 25 ~ "Below Average",
    TRUE ~ "Poor"
  )
}

# Apply tiers for each archetype
intl_per_game <- intl_per_game %>%
  mutate(
    scorer_tier = get_tier(scorer_percentile),
    post_scorer_tier = get_tier(post_scorer_percentile),
    rebounder_tier = get_tier(rebounder_percentile),
    facilitator_tier = get_tier(facilitator_percentile),
    shot_blocker_tier = get_tier(shot_blocker_percentile),
    floor_spacer_tier = get_tier(floor_spacer_percentile)
  )

# Create a summary dataframe with core player info + percentiles
player_archetypes <- intl_per_game %>%
  dplyr::select(first_name, last_name, age, nba_experience,
         scorer_percentile, post_scorer_percentile, rebounder_percentile,
         facilitator_percentile, shot_blocker_percentile, floor_spacer_percentile,
         scorer_tier, post_scorer_tier, rebounder_tier,
         facilitator_tier, shot_blocker_tier, floor_spacer_tier, games)

# Compute each player's top 3 archetypes by percentile
player_archetypes <- player_archetypes %>%
  rowwise() %>%
  mutate(
    # Get the archetype names and scores
    archetypes = list(c("Scorer", "Post Scorer", "Rebounder",
                        "Facilitator", "Shot Blocker", "Floor Spacer")),
    scores = list(c(scorer_percentile, post_scorer_percentile, rebounder_percentile,
                    facilitator_percentile, shot_blocker_percentile, floor_spacer_percentile)),

    # Order top 3 by percentile
    top3_indices = list(order(unlist(scores), decreasing = TRUE)[1:3]),
    top3_archetypes = list(unlist(archetypes)[unlist(top3_indices)]),
    top3_scores = list(unlist(scores)[unlist(top3_indices)]),

    # Extract top 3 + labels
    archetype_1 = top3_archetypes[[1]],
    archetype_2 = top3_archetypes[[2]],
    archetype_3 = top3_archetypes[[3]],
    archetype_1_label = paste(get_tier(top3_scores[[1]]), archetype_1),
    archetype_2_label = paste(get_tier(top3_scores[[2]]), archetype_2),
    archetype_3_label = paste(get_tier(top3_scores[[3]]), archetype_3)
  ) %>%
  ungroup() %>%
  dplyr::select(first_name, last_name, age, nba_experience,
         archetype_1_label, archetype_2_label, archetype_3_label,
         scorer_percentile:floor_spacer_percentile)
player_archetypes

```





# Capable Bigs: Filter, 3D Plot, Top-3 Table
```{r}
# ---- 1) Filter & prep ----
capable_bigs <- intl_per_game %>%
  filter(
    games >= 10,
    minutes >=15, 
    age <= 33,
    most_recent_year >= 2019,
    !is.na(post_scorer_percentile),
    !is.na(shot_blocker_percentile),
    !is.na(rebounder_percentile),
    !is.na(nba_probability),
    nba_probability > 0.5,              
    nba_experience == FALSE
  ) %>%
  mutate(
    balance_score = (post_scorer_percentile + shot_blocker_percentile + rebounder_percentile) / 3
  )

# ---- 2) Identify top 3 by overall big score ----
top_n <- min(3, nrow(capable_bigs))
top3_idx <- order(capable_bigs$balance_score, decreasing = TRUE)[seq_len(top_n)]

# Colors: top 3 = orange, others = blue
cols <- rep("#0072B2", nrow(capable_bigs))
if (top_n > 0) cols[top3_idx] <- "#D55E00"

# ---- 3) 3D plot ----
s3d <- scatterplot3d(
  x = capable_bigs$post_scorer_percentile,
  y = capable_bigs$shot_blocker_percentile,
  z = capable_bigs$rebounder_percentile,
  xlab = "Post Scorer Percentile (%)",
  ylab = "Shot Blocker Percentile (%)",
  zlab = "Rebounder Percentile (%)",
  color = cols,
  pch = 19,
  cex.symbols = 0.9,
  main = "Capable Bigs (Post Scoring, Rim Protection, Rebounding)",
  angle = 55 
)

# Re-plot and label top 3
if (top_n > 0) {
  s3d$points3d(
    capable_bigs$post_scorer_percentile[top3_idx],
    capable_bigs$shot_blocker_percentile[top3_idx],
    capable_bigs$rebounder_percentile[top3_idx],
    col = "#D55E00", pch = 19, cex = 1.15
  )

  lab <- s3d$xyz.convert(
    capable_bigs$post_scorer_percentile[top3_idx],
    capable_bigs$shot_blocker_percentile[top3_idx],
    capable_bigs$rebounder_percentile[top3_idx]
  )

  text(
    lab$x, lab$y,
    labels = paste(capable_bigs$first_name[top3_idx], capable_bigs$last_name[top3_idx]),
    pos = 2, cex = 0.7, font = 2, col = "#D55E00"
  )
}

capable_bigs_top3 <- capable_bigs %>%
  arrange(desc(balance_score)) %>%             
  slice_head(n = 3) %>%                         # top 3 only
  mutate(
    rebounds = coalesce(
      offensive_rebounds + defensive_rebounds
    )
  ) %>%
  dplyr::select(
    first_name, last_name,
    points, rebounds, assists,                 # box score
    games, age, minutes,                               # context
    scorer_tier, post_scorer_tier, rebounder_tier,
    facilitator_tier, shot_blocker_tier, floor_spacer_tier
  )

print(capable_bigs_top3, n = 3)

```




# Lead Guards: Filter, 3D Plot
```{r}
# ---- 1) Filter & prep data ----
guards_3d <- intl_per_game %>%
  filter(
    games >= 10,
    age <= 33,
    minutes >= 15, 
    nba_probability > 0.5,
    most_recent_year >= 2019,
    nba_experience %in% c(FALSE, 0),       # exclude NBA players
    !is.na(scorer_percentile),
    !is.na(facilitator_percentile),
    !is.na(nba_probability)
  ) %>%
  mutate(
    guard_score = (scorer_percentile + facilitator_percentile) / 2,
    nba_prob_pct = 100 * nba_probability
  )

# ---- 2) Identify top 5 ----
top5_idx <- order(guards_3d$guard_score, decreasing = TRUE)[1:3]

# ---- 3) Assign colors ----
cols <- rep("#0072B2", nrow(guards_3d))
cols[top5_idx] <- "#D55E00"

# ---- 4) 3D Plot ----
s3d <- scatterplot3d(
  x = guards_3d$scorer_percentile,
  y = guards_3d$facilitator_percentile,
  z = guards_3d$nba_prob_pct,
  xlab = "Scorer Percentile (%)",
  ylab = "Facilitator Percentile (%)",
  zlab = "NBA Probability (%)",
  color = cols,
  pch = 19,
  cex.symbols = 0.9,
  main = "Lead Guards (Scoring, Facilitation, NBA Probability)"
)

# ---- 5) Highlight + label top 5 ----
if (length(top5_idx) > 0) {
  s3d$points3d(
    guards_3d$scorer_percentile[top5_idx],
    guards_3d$facilitator_percentile[top5_idx],
    guards_3d$nba_prob_pct[top5_idx],
    col = "#D55E00", pch = 19, cex = 1.1
  )

  lab <- s3d$xyz.convert(
    guards_3d$scorer_percentile[top5_idx],
    guards_3d$facilitator_percentile[top5_idx],
    guards_3d$nba_prob_pct[top5_idx]
  )

  text(
    lab$x, lab$y,
    labels = paste(guards_3d$first_name[top5_idx], guards_3d$last_name[top5_idx]),
    pos = 4,
    cex = 0.7,
    font = 2,
    col = "#D55E00"
  )
}


```
```{r}
guards_top3 <- guards_3d %>%
  mutate(rank_guard = rank(-guard_score, ties.method = "first")) %>%
  filter(rank_guard <= 3) %>%
  arrange(rank_guard) %>%
  mutate(
    Name = paste(first_name, last_name),
    # If you have total_rebounds, use that; otherwise sum O/D boards
    REB = dplyr::coalesce(offensive_rebounds + defensive_rebounds),
    PTS = points,
    AST = assists,
    G   = games
  ) %>%
  dplyr::select(Name, age, PTS, REB, AST, G,
         scorer_tier, facilitator_tier, floor_spacer_tier)

print(guards_top3, n = 3)
```


# 3&D Wings: Build Defense Metric, Filter, 3D Plot
```{r}
# --- 1) Build Defense Percentile (Steal% + Block%) ---
intl_per_game <- intl_per_game %>%
  mutate(
    steal_pctile = percent_rank(steal_percentage_wa),
    block_pctile = percent_rank(block_percentage_wa),
    defense_score = (steal_pctile + block_pctile) / 2,
    defense_percentile = 100 * round(percent_rank(defense_score), 3)
  )

# --- 2) Filter data for 3&D view ---
three_d_wings <- intl_per_game %>%
  filter(
    games >= 10,
    minutes >= 15,
    age <= 33,
    nba_probability >= 0.5, 
    nba_experience == FALSE, 
    most_recent_year >= 2019,
    !is.na(floor_spacer_percentile),
    !is.na(defense_percentile),
    !is.na(nba_probability)
  ) %>%
  mutate(
    nba_prob_pct = 100 * nba_probability,
    three_d_combo = (floor_spacer_percentile + defense_percentile) / 2
  )

# --- 3) Identify top 3 players by 3&D combo score ---
top3_idx <- order(three_d_wings$three_d_combo, decreasing = TRUE)[1:3]

# Assign colors: top 3 orange, rest blue
cols <- rep("#0072B2", nrow(three_d_wings))
cols[top3_idx] <- "#D55E00"

# --- 4) 3D Plot ---
s3d <- scatterplot3d(
  x = three_d_wings$floor_spacer_percentile,
  y = three_d_wings$defense_percentile,
  z = three_d_wings$nba_prob_pct,
  xlab = "Floor Spacer Percentile (%)",
  ylab = "Defense Percentile (%)",
  zlab = "NBA Probability (%)",
  color = cols,
  pch = 19,
  cex.symbols = 0.9,
  main = "3&D Wings (Shooting, Defense, NBA Probability)"
)

# --- 5) Highlight and label top 3 ---
if (length(top3_idx) > 0) {
  s3d$points3d(
    three_d_wings$floor_spacer_percentile[top3_idx],
    three_d_wings$defense_percentile[top3_idx],
    three_d_wings$nba_prob_pct[top3_idx],
    col = "#D55E00", pch = 19, cex = 1.15
  )

  lab <- s3d$xyz.convert(
    three_d_wings$floor_spacer_percentile[top3_idx],
    three_d_wings$defense_percentile[top3_idx],
    three_d_wings$nba_prob_pct[top3_idx]
  )
  text(
    lab$x, lab$y,
    labels = paste(three_d_wings$first_name[top3_idx], three_d_wings$last_name[top3_idx]),
    pos = 4, cex = 0.7, font = 2, col = "#D55E00"
  )
}


```
```{r}
three_d_top3 <- three_d_wings %>%
  arrange(desc(three_d_combo)) %>%              # rank by 3&D composite
  slice_head(n = 3) %>%                         # top 3 only
  mutate(
    Name = paste(first_name, last_name),
    PTS  = round(points, 1),
    REB  = round(coalesce(offensive_rebounds + defensive_rebounds), 1),
    AST  = round(assists, 1),
    G    = games,
    defense_tier = if ("defense_tier" %in% names(.)) defense_tier else get_tier(defense_percentile)
  ) %>%
  dplyr::select(Name, age, PTS,  AST, G,
         floor_spacer_tier, defense_tier, three_point_percentage)

print(three_d_top3, n = 3)

```

# Two-Way Guards: Filter, 3D Plot
```{r}

# --- 1) Filter + two-way guard score ---
two_way_guards <- intl_per_game %>%
  filter(
    games >= 10,
    age <= 33,
    nba_experience == FALSE, 
    nba_probability >= 0.5,
    most_recent_year >=2019, 
    minutes >= 15,
    !is.na(scorer_percentile),
    !is.na(defense_percentile),
    !is.na(nba_probability)
  ) %>%
  mutate(
    nba_prob_pct = 100 * nba_probability,
    two_way_guard_score = (scorer_percentile + defense_percentile) / 2
  )

# --- 2) Top 3 by two-way score ---
top3_idx <- order(two_way_guards$two_way_guard_score, decreasing = TRUE)[1:3]
cols <- rep("#0072B2", nrow(two_way_guards))
cols[top3_idx] <- "#D55E00"

# --- 3) 3D plot: Scoring vs Defense vs NBA Prob ---
s3d <- scatterplot3d(
  x = two_way_guards$scorer_percentile,
  y = two_way_guards$defense_percentile,
  z = two_way_guards$nba_prob_pct,
  xlab = "Scorer Percentile",
  ylab = "Defense Percentile (Steal% + Block%)",
  zlab = "NBA Probability (%)",
  color = cols,
  pch = 19,
  cex.symbols = 0.9,
  main = "3D: Two-Way Guards (Scoring, Defense, NBA Probability)"
)

# highlight + label top 3
s3d$points3d(
  two_way_guards$scorer_percentile[top3_idx],
  two_way_guards$defense_percentile[top3_idx],
  two_way_guards$nba_prob_pct[top3_idx],
  col = "#D55E00", pch = 19, cex = 1.15
)
lab <- s3d$xyz.convert(
  two_way_guards$scorer_percentile[top3_idx],
  two_way_guards$defense_percentile[top3_idx],
  two_way_guards$nba_prob_pct[top3_idx]
)
text(
  lab$x, lab$y,
  labels = paste(two_way_guards$first_name[top3_idx], two_way_guards$last_name[top3_idx]),
  pos = 4, cex = 0.7, font = 2, col = "#D55E00"
)


```

# Two-Way Guards: Ranked Table
```{r}
# --- 4) Table: top 3 first, then everyone else ---
two_way_guards %>%
  mutate(is_top3 = row_number() %in% top3_idx) %>%
  arrange(desc(is_top3), desc(two_way_guard_score)) %>%
  dplyr::select(first_name, last_name, age, minutes, games,
         scorer_percentile, defense_percentile, nba_probability,
         two_way_guard_score, is_top3) %>%
  print(n = 20)

```

# Archetype Translation: % with NBA Experience (Top 30%)
```{r}
# --- 1) Gather all archetype percentiles ---
arche_cols <- c(
  "scorer_percentile",
  "facilitator_percentile",
  "floor_spacer_percentile",
  "rebounder_percentile",
  "shot_blocker_percentile",
  "post_scorer_percentile",
  "defense_percentile"
)

arche_long <- intl_per_game %>%
  filter(!is.na(nba_experience)) %>%
  dplyr::select(first_name, last_name, nba_experience, all_of(arche_cols)) %>%
  pivot_longer(
    cols = all_of(arche_cols),
    names_to = "archetype",
    values_to = "percentile"
  ) %>%
  mutate(
    archetype = recode(
      archetype,
      scorer_percentile = "Scorer",
      facilitator_percentile = "Facilitator",
      floor_spacer_percentile = "Floor Spacer",
      rebounder_percentile = "Rebounder",
      shot_blocker_percentile = "Shot Blocker",
      post_scorer_percentile = "Post Scorer",
      defense_percentile = "Defense"
    )
  )

# --- 2) For each archetype, count how many players with NBA experience are Good-or-better ---
arche_value <- arche_long %>%
  filter(percentile >= 70) %>%
  group_by(archetype) %>%
  summarise(
    total_players = n(),
    nba_players = sum(nba_experience == TRUE, na.rm = TRUE),
    pct_nba = nba_players / total_players * 100
  ) %>%
  arrange(desc(pct_nba))

print(arche_value)

# --- 3) Plot NBA representation by archetype ---
ggplot(arche_value, aes(x = reorder(archetype, pct_nba), y = pct_nba)) +
  geom_col(fill = "steelblue", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", pct_nba)), vjust = -0.3, size = 3.8) +
  labs(
    title = "NBA Representation by Archetype (Top 30% Performers)",
    x = NULL,
    y = "Percent with NBA Experience"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())

```
```{r}
nba_exp_summary <- intl_per_game %>%
  summarise(
    total_players = n(),
    nba_players = sum(nba_experience == TRUE, na.rm = TRUE),
    pct_nba = round(100 * nba_players / total_players, 2)
  )

```

# Shiny App Source (Not Evaluated)
```{r, eval=FALSE, echo=TRUE}

# --- assume intl_per_game already exists with percentiles, *_tier, nba_probability ---
# Build/ensure a single consistent defense metric: defense_percentile
players <- intl_per_game %>%
  mutate(
    defense_percentile = if ("defense_percentile" %in% names(intl_per_game)) {
      defense_percentile
    } else {
      # Fallback: average of block and (optional) steal percentiles if present
      rowMeans(cbind(
        shot_blocker_percentile,
        if ("steal_percentile" %in% names(intl_per_game)) steal_percentile else NA_real_
      ), na.rm = TRUE)
    }
  ) %>%
  dplyr::select(
    first_name, last_name, age, nba_experience, most_recent_year,
    games, minutes, nba_probability,
    scorer_percentile, post_scorer_percentile, rebounder_percentile,
    facilitator_percentile, shot_blocker_percentile, floor_spacer_percentile,
    defense_percentile,
    scorer_tier, post_scorer_tier, rebounder_tier,
    facilitator_tier, shot_blocker_tier, floor_spacer_tier,
    dplyr::any_of(c("currently_in_nba"))  # optional
  )

# Dropdown mapping for 3D axis choices
percentile_cols <- c(
  "Scorer"        = "scorer_percentile",
  "Post Scorer"   = "post_scorer_percentile",
  "Rebounder"     = "rebounder_percentile",
  "Facilitator"   = "facilitator_percentile",
  "Rim Protector" = "shot_blocker_percentile",
  "Floor Spacer"  = "floor_spacer_percentile",
  "Defensive"     = "defense_percentile"
)

ui <- fluidPage(
  titlePanel("Kings Player Finder"),
  sidebarLayout(
    sidebarPanel(
      textInput("q", "Search name", ""),
      sliderInput("age_max","Max age", min = 16, max = 45, value = 30, step = 1),
      sliderInput("min_games","Min games", min = 0, max = 82, value = 10, step = 1),
      sliderInput("min_minutes","Min minutes", min = 0, max = 40, value = 15, step = 1),

      radioButtons(
        "nba_status", "NBA status",
        choices = c("Any" = "any", "NBA only" = "nba_only", "Not currently in NBA" = "not_nba"),
        selected = "any", inline = TRUE
      ),
      checkboxInput("nba_ready","NBA ready (nba_probability > 0.5)", FALSE),

      tags$hr(),
      h4("Archetype filters (percentile mins)"),
      sliderInput("min_scorer", "Scorer",         0, 100, 0, step = 5),
      sliderInput("min_post",   "Post Scorer",    0, 100, 0, step = 5),
      sliderInput("min_reb",    "Rebounder",      0, 100, 0, step = 5),
      sliderInput("min_fac",    "Facilitator",    0, 100, 0, step = 5),
      sliderInput("min_block",  "Rim Protector",  0, 100, 0, step = 5),
      sliderInput("min_space",  "Floor Spacer",   0, 100, 0, step = 5),
      sliderInput("min_def",    "Defense",        0, 100, 0, step = 5),

      tags$hr(),
      h4("Quick presets"),
      actionButton("preset_big","Two-way Big (post+block+reb 70+)"),
      actionButton("preset_guard","Lead Guard (facil+scorer 85+)"),
      actionButton("preset_3nd","3-and-D Wing (space 80+, block 60+)"),

      tags$hr(),
      downloadButton("dl", "Download CSV")
    ),
    mainPanel(
      tabsetPanel(
        tabPanel("Finder (Table)", DTOutput("tbl")),
        tabPanel("3D Graph",
          fluidRow(
            column(4, selectInput("x_var", "X axis", choices = percentile_cols, selected = "post_scorer_percentile")),
            column(4, selectInput("y_var", "Y axis", choices = percentile_cols, selected = "shot_blocker_percentile")),
            column(4, selectInput("z_var", "Z axis", choices = percentile_cols, selected = "rebounder_percentile"))
          ),
          plotlyOutput("p3d", height = "620px"),
          helpText("Top 3 by the mean of selected axes are highlighted in orange and labeled. Only players with most_recent_year >= 2019 are shown.")
        )
      )
    )
  )
)

server <- function(input, output, session){

  # Presets
  observeEvent(input$preset_big, {
    updateSliderInput(session,"min_post",  value = 70)
    updateSliderInput(session,"min_block", value = 70)
    updateSliderInput(session,"min_reb",   value = 70)
  })
  observeEvent(input$preset_guard, {
    updateSliderInput(session,"min_fac",   value = 85)
    updateSliderInput(session,"min_scorer",value = 85)
  })
  observeEvent(input$preset_3nd, {
    updateSliderInput(session,"min_space", value = 80)
    updateSliderInput(session,"min_block", value = 60)
  })

  # Core filtered data
  filtered <- reactive({
    df <- players %>%
      filter(!is.na(most_recent_year) & most_recent_year >= 2019) %>%
      filter(
        is.na(age) | age <= input$age_max,
        games >= input$min_games,
        minutes >= input$min_minutes,
        scorer_percentile       >= input$min_scorer,
        post_scorer_percentile  >= input$min_post,
        rebounder_percentile    >= input$min_reb,
        facilitator_percentile  >= input$min_fac,
        shot_blocker_percentile >= input$min_block,
        floor_spacer_percentile >= input$min_space,
        defense_percentile      >= input$min_def
      )

    # NBA status
    if (input$nba_status == "nba_only") {
      df <- df %>% filter(if ("currently_in_nba" %in% names(.)) currently_in_nba else nba_experience)
    } else if (input$nba_status == "not_nba") {
      df <- df %>% filter(if ("currently_in_nba" %in% names(.)) !currently_in_nba else !nba_experience)
    }

    # NBA ready threshold
    if (input$nba_ready) df <- df %>% filter(!is.na(nba_probability) & nba_probability > 0.5)

    # Name search
    if (nzchar(input$q)) {
      pat <- tolower(input$q)
      df <- df %>%
        mutate(full_name = paste(first_name, last_name)) %>%
        filter(grepl(pat, tolower(full_name), fixed = TRUE)) %>%
        select(-full_name)
    }

    df %>%
      mutate(
        big_balance = rowMeans(cbind(post_scorer_percentile,
                                     shot_blocker_percentile,
                                     rebounder_percentile), na.rm = TRUE),
        guard_score = rowMeans(cbind(facilitator_percentile,
                                     scorer_percentile), na.rm = TRUE)
      ) %>%
      arrange(desc(pmax(big_balance, guard_score, scorer_percentile)))
  })

  output$tbl <- renderDT({
    datatable(
      filtered(),
      options = list(pageLength = 25, scrollX = TRUE),
      rownames = FALSE
    )
  })

  # 3D plot with selectable axes & top-3 highlight
  output$p3d <- renderPlotly({
    df <- filtered()
    req(nrow(df) > 0)

    xcol <- input$x_var; ycol <- input$y_var; zcol <- input$z_var

    df <- df %>%
      mutate(
        axes_mean = rowMeans(cbind(.data[[xcol]], .data[[ycol]], .data[[zcol]]), na.rm = TRUE)
      ) %>%
      arrange(desc(axes_mean)) %>%
      mutate(
        rank_axes = row_number(),
        is_top3   = rank_axes <= 3
      )

    has_prob <- "nba_probability" %in% names(df)

    df <- df %>%
      mutate(
        label = paste0(
          first_name, " ", last_name,
          "<br>Age: ", age,
          " | G: ", games, " | Min: ", minutes,
          "<br>", names(percentile_cols)[percentile_cols == xcol], ": ", round(.data[[xcol]]),
          "<br>", names(percentile_cols)[percentile_cols == ycol], ": ", round(.data[[ycol]]),
          "<br>", names(percentile_cols)[percentile_cols == zcol], ": ", round(.data[[zcol]]),
          if (has_prob) paste0("<br>NBA Prob: ", sprintf("%.2f", nba_probability)) else ""
        )
      )

    df_top  <- df %>% filter(is_top3)
    df_rest <- df %>% filter(!is_top3)

    p <- plot_ly(type = "scatter3d", mode = "markers")

    if (nrow(df_rest) > 0) {
      p <- add_trace(
        p, data = df_rest,
        x = ~.data[[xcol]], y = ~.data[[ycol]], z = ~.data[[zcol]],
        text = ~label, hoverinfo = "text",
        marker = list(size = 4),
        name = "Others", showlegend = TRUE
      )
    }

    if (nrow(df_top) > 0) {
      p <- add_trace(
        p, data = df_top,
        x = ~.data[[xcol]], y = ~.data[[ycol]], z = ~.data[[zcol]],
        text = ~label, hoverinfo = "text",
        marker = list(size = 7, color = "orange", line = list(width = 1)),
        name = "Top 3 (by selected axes)", showlegend = TRUE
      ) %>%
      add_text(
        data = df_top,
        x = ~.data[[xcol]], y = ~.data[[ycol]], z = ~.data[[zcol]],
        text = ~paste(first_name, last_name),
        textposition = "top center",
        showlegend = FALSE
      )
    }

    p %>% layout(
      scene = list(
        xaxis = list(title = names(percentile_cols)[percentile_cols == xcol]),
        yaxis = list(title = names(percentile_cols)[percentile_cols == ycol]),
        zaxis = list(title = names(percentile_cols)[percentile_cols == zcol])
      )
    )
  })

  output$dl <- downloadHandler(
    filename = function() paste0("kings_player_finder_", Sys.Date(), ".csv"),
    content = function(file) write.csv(filtered(), file, row.names = FALSE)
  )
}

shinyApp(ui, server)

```



